Demes: a specification for interchanging complex demography

### 2022-04-05 paper.tex
Update paper.tex
Co-authored-by: Jerome Kelleher <jk@well.ox.ac.uk>

% local definitions
\newcommand{\ms}[0]{\texttt{ms}}
\newcommand{\msprime}[0]{\texttt{msprime}}
\newcommand{\stdpopsim}[0]{\texttt{stdpopsim}}
\newcommand{\demes}[0]{\texttt{demes}}
\newcommand{\demesdraw}[0]{\texttt{demesdraw}}
\newcommand{\Demes}[0]{\texttt{Demes}}
\newcommand{\moments}[0]{\texttt{moments}}
\newcommand{\dadi}[0]{\texttt{$\partial$a$\partial$i}}
\newcommand{\fwdpy}[0]{\texttt{fwdpy11}}
\newcommand{\slim}[0]{\texttt{SLiM}}
\newcommand{\gadma}[0]{\texttt{GADMA}}

\newcommand{\aprcomment}[1]{{\textcolor{blue}{APR: #1}}}
\newcommand{\jkcomment}[1]{{\textcolor{red}{JK: #1}}}
\newcommand{\ggcomment}[1]{{\textcolor{yellow!60!red}{GG: #1}}}
\newcommand{\krtcomment}[1]{{\textcolor{purple}{KRT: #1}}}
\newcommand{\mhcomment}[1]{{\textcolor{cyan}{MH: #1}}}

@@

Inference methods for demography usually output a set of
text files which describe the inferred model,
and nearly all methods use different file formats.
Inference results are typically reported in publications
via a combination of visual depiction,
a list of key parameters in tabular form and a discussion within the text.
Unfortunately these descriptions are often ambiguous, and
implementing the precise model inferred for later simulation
is at best tedious and error prone~\citep{ragsdale2020lessons}
and occasionally impossible because of missing information. 
\mhcomment{Maybe add a short, concrete example to convince readers of the problem?}
%\aprcomment{This feels tangential - maybe discussion}
%The \stdpopsim\ project is a community-maintained catalog of species
%information and demographic models designed to provide standardised
%population genetic simulations~\citep{adrion2020community}.
%The development process for adding demographic models to
%\stdpopsim\ tries to avoid errors in model implementation by requiring two
%independent implementations of each model
%to be equivalent. While this process is
%robust and has uncovered both significant
%errors~\citep{ragsdale2020lessons} and ambiguities in
%published models, it is labour intensive.
Instead, it would be far better if published results were reported
in an unambiguous standardised format that could be provided
directly as input to simulators.

@@

Here we present ``Demes'', a data model and file format specification for
complex demographic models developed by the
PopSim Consortium~\citep{adrion2020community}.
The Demes data model precisely defines the sizes and relationships
of populations, and it provides a way to explicitly encode the
information relevant to demography while avoiding repetition.
This data model is implemented in the widely used YAML
format~\citep{ben2009yaml}, which provides a good balance between human
and machine readability.
The specification precisely defines the required behaviour of implementations,
ensuring that there is no ambiguity of interpretation, and includes both a
reference implementation and an extensive suite of test examples and their
expected output.
The initial software ecosystem includes high-quality Python
and C parser implementations, as well as utilities for verification and
visualisation of Demes models, and has been implemented in several popular
inference and simulation methods (Table~\ref{tab:software}).
We hope that this data model and file format will be widely adopted
by the community, such that users can expect to simulate directly
from inferred models with little programming effort.

@@

@@

\krtcomment{We need to revisit the paragraph(s) below.  We do not assume individuals are exchangeable, because we do not
	require that all individuals have the same fitness. See, e.g., p 60 of Wakeley for why fitness == exchangeability
	and why that is required for the standard coalescent.}
\mhcomment{The `exchangeability' assumption also breaks down under 
	different reproductive modes, since the sampling of the alleles affects coalescent times. 
	For example, in a highly selfing species the coalescent time for two different alleles 
	within an individual is much lower than two alleles from different individuals. I've tried 
	to rewrite the below paragraph to update and clarify the reproduction parameters in 
	line with the discussions on Github. Finally, 'rates' imply events occur as Poisson-like 
	processes over continuous time, whereas the values used here instead define the 
	\emph{fraction} of reproductions of each type. Shall we rename to selfing/clonal `fractions',
	 or would that mess up the rest of the schema too much?}

Each population has a selfing rate and cloning rate assigned to it, where each denotes 
the probability that offspring are generated from one generation to the next by either 
self-fertilisation or cloning of an individual. More specifically, denote the clonal rate by 
$\sigma$ and the selfing rate by $S$. In each population a proportion of offspring 
$\sigma$ reproduce clonally, while $1-\sigma$ reproduce sexually. Within the 
sexually--reproducing offspring, $S$ were born via self--fertilisation while the rest were 
outcrossed. Hence $S$ and $\sigma$ can take any value between zero and one, but 
do not have to sum to one. See \citet{hartfield2016facsexcoal} for more details of the 
mathematical properties of these rates in a coalescent context. \mhcomment{If requested 
I can quickly draw up a schematic diagram to illustrate these partitions.} Note that even if 
$S$, $\sigma$ are set to zero then some residual selfing can occur due to sampling in 
the Wright--Fisher model.

We omit any details regarding selection.
Thus, the history of population size changes does not depend on mean fitnesses within
a population nor on differences in mean fitness between populations.
This indepedence of population size and
selection is often called ``soft" selection \citep{christiansen1975hard}.

>>>>

% local definitions
\newcommand{\ms}[0]{\texttt{ms}}
\newcommand{\msprime}[0]{\texttt{msprime}}
\newcommand{\stdpopsim}[0]{\texttt{stdpopsim}}
\newcommand{\demes}[0]{\texttt{demes}}
\newcommand{\demesdraw}[0]{\texttt{demesdraw}}
\newcommand{\Demes}[0]{\texttt{Demes}}
\newcommand{\moments}[0]{\texttt{moments}}
\newcommand{\dadi}[0]{\texttt{$\partial$a$\partial$i}}
\newcommand{\fwdpy}[0]{\texttt{fwdpy11}}
\newcommand{\slim}[0]{\texttt{SLiM}}
\newcommand{\gadma}[0]{\texttt{GADMA}}
\newcommand{\tskit}[0]{\texttt{tskit}}

\newcommand{\aprcomment}[1]{{\textcolor{blue}{APR: #1}}}
\newcommand{\jkcomment}[1]{{\textcolor{red}{JK: #1}}}
\newcommand{\ggcomment}[1]{{\textcolor{yellow!60!red}{GG: #1}}}
\newcommand{\krtcomment}[1]{{\textcolor{purple}{KRT: #1}}}

@@

Inference methods for demography usually output a set of
text files which describe the inferred model,
and nearly all methods use different file formats.
Inference results are typically reported in publications
via a combination of visual depiction,
a list of key parameters in tabular form and a discussion within the text.
Unfortunately these descriptions are often ambiguous, and
implementing the precise model inferred for later simulation
is at best tedious and error prone~\citep{ragsdale2020lessons}
and occasionally impossible because of missing information.
%\aprcomment{This feels tangential - maybe discussion}
%The \stdpopsim\ project is a community-maintained catalog of species
%information and demographic models designed to provide standardised
%population genetic simulations~\citep{adrion2020community}.
%The development process for adding demographic models to
%\stdpopsim\ tries to avoid errors in model implementation by requiring two
%independent implementations of each model
%to be equivalent. While this process is
%robust and has uncovered both significant
%errors~\citep{ragsdale2020lessons} and ambiguities in
%published models, it is labour intensive.
Instead, it would be far better if published results were reported
in an unambiguous standardised format that could be provided
directly as input to simulators.

@@

Here we present ``Demes'', a data model and file format specification for
complex demographic models developed by the
PopSim Consortium~\citep{adrion2020community}.
The Demes data model precisely defines the sizes and relationships
of populations, and it provides a way to explicitly encode the
information relevant to demography while avoiding repetition.
This data model is implemented in the widely used YAML
format~\citep{ben2009yaml}, which provides a good balance between human
and machine readability.
The specification precisely defines the required behaviour of implementations,
ensuring that there is no ambiguity of interpretation, and includes both a
reference implementation and an extensive suite of test examples and their
expected output.
The initial software ecosystem includes high-quality Python
and C parser implementations, as well as utilities for verification and
visualisation of Demes models, and has been implemented in several popular
inference and simulation methods (Table~\ref{tab:software}).
We hope that this data model and file format will be widely adopted
by the community, such that users can expect to simulate directly
from inferred models with little or no programming effort.

@@

Here, we highlight the interface between Demes and other software, including
simulation and model illustration tools. Demes allows us to specify a
demographic model which can be used as the input for a growing
number of simulation software (Table~\ref{tab:software}). We implemented the
human two-population demographic model from \citet{tennessen2012evolution}
inferred from European and African-American sequencing data. This model (given
in Demes format in Figure~\ref{fig:tennessen}) is parameterized by an ancestral
population with an ancient growth, divergence into ``AFR'' and ``EUR'' that
each have multiple-epoch size histories, and multiple epochs of continuous
migration between the two branches (illustrated using \demesdraw\ in
Figure~\ref{fig:showcase}A). The large final sizes (\(\approx 500,000\)
individuals each) are one to three orders of magnitude larger than ancestral
population sizes, reflecting the recent explosive population size increase in
human populations.

We used this model to simulate 20 haploid genome copies from each EUR and AFR
at time zero (i.e., present day) to obtain the joint site-frequency spectrum
(SFS), a summary of observed allele frequencies widely used in evolutionary
inference
\citep{bustamante2001directional,gutenkunst2009inferring,tennessen2012evolution,jouganous2017inferring,kamm2017efficient,kim2017inference}.
We used \msprime\ \citep{baumdicker2021-iu} to simulate a large recombining
region with the mutation rate assumed in \citet{tennessen2012evolution} and
computed the observed SFS using \tskit\ \citep{ralph2020efficiently}. We
compared that simulated SFS to its expectation computed using \moments\
\citep{jouganous2017inferring} (Figure~\ref{fig:showcase}B,C). For both
\msprime\ and \moments\ simulations, the same demographic model was used as
input, avoiding the need to specify the complex demographic model multiple
times using the unique API of each software.

@@

\krtcomment{We need to revisit the paragraph(s) below.  We do not assume individuals are exchangeable, because we do not
	require that all individuals have the same fitness. See, e.g., p 60 of Wakeley for why fitness == exchangeability
	and why that is required for the standard coalescent.}

Within a population, individuals are assumed to be exchangeable, but there are
also parameters for nonrandom mating, such as selfing or cloning rates, which
give the probability that offspring are generated from one generation to the
next by self-fertilisation or cloning of an individual. Cloning and selfing
rates can take any value between zero and one, and if both are specified then
the selfing rate is the probability an offspring is generated through
self-fertilisation given it was produced sexually (i.e. with probability one
minus the cloning rate). We omit any details regarding selection.
Thus, the history of population size changes does not depend on mean fitnesses within
a population nor on differences in mean fitness between populations.
This indepedence of population size and
selection is often called ``soft" selection \citep{christiansen1975hard}.
